---
title: "IreneChang_A02_CodingBasics.Rmd"
author: "Irene Chang"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: inline
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Sakai.

## Basics Day 1

1.  Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name.

2.  Compute the mean and median of this sequence.

3.  Ask R to determine whether the mean is greater than the median.

4.  Insert comments in your code to describe what you are doing.

```{r}
#1. 
seq1 <- seq(1,100,4) #from, to, by and also naming the sequence "seq1"

#2. 
mean(seq1) #finding the mean of the sequence that I named
median(seq1) #finding the median of the sequence that I named

#3. 
mean(seq1)>median(seq1) #this is showing if the mean of the sequence is greater than the median, so use the conditional statement

```

## Basics Day 2

5.  Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6.  Label each vector with a comment on what type of vector it is.

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

8.  Label the columns of your data frame with informative titles.

```{r}
name <- c('Ellen', 'Dan', 'Jane', 'Irene') #character
scores <- c(99, 48, 82, 95) #numeric
pass <- ifelse(scores>=50, T, F) #logical

class_scores <- data.frame (name=name, scores=scores, pass=pass)

class_scores


```

9.  QUESTION: How is this data frame different from a matrix?

> Answer: A matrix can only contain one class, whereas a data frame can have many different classes of data.

10. Create a function with an if/else statement. Your function should take a **vector** of test scores and print (not return) whether a given test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. 

11. Apply your function to the vector with test scores that you created in number 5.

```{r}
PassExam <- function(scores) {
  reviewscores <- ifelse(scores>=50, T, F)
  print(reviewscores)
}
PassExam(scores)

# PassExam <- function(scores) {
#   if(scores < 50) {
#     FALSE
#   }
#   else {
#     TRUE
#   }
# }

#PassExam(scores)
```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: 'ifelse' works but 'if' and 'else' does not work because it can't run through a vector of a length greater than 1. The error that I get is "Error in if (scores < 50) { : the condition has length > 1". Since the vector that we are using has multiple elements, the 'if' and 'else' function does not work because it is only expecting one.
