---
title: "IreneChang_A06_GLMs.Rmd"
author: "Irene Chang"
date: "Spring 2023"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A06_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup2}
#1
getwd()
library(tidyverse)
library(agricolae)
library(here)
library(lubridate)

here()

NTL.Lake <- read.csv(here("Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv"), stringsAsFactors = TRUE)
NTL.Lake$sampledate <- mdy(NTL.Lake$sampledate)

#2
mytheme <-
  theme_classic() +
  theme(axis.text = element_text(color = "black", hjust = 0.5),
        plot.title = element_text(hjust = 0.5),
        legend.position = "bottom")
theme_set(mytheme)

```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: The mean lake temperature recorded during July does not change with the depth across all lakes.
Ha: The mean lake temperature recorded during July does change with the depth across all lakes.


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4

NTL.LTER.month <- mutate(NTL.Lake, month = month(sampledate))

NTL.LTER.July <- NTL.LTER.month %>% 
  filter(month == 7)

NTL.LTER.subset <- NTL.LTER.July %>% 
  select(lakename, year4, daynum, depth, temperature_C) %>% 
  na.omit(lakename, year4, daynum, depth, temperature_C)

#5

NTL.tempbydepth <-
  ggplot(NTL.LTER.subset, aes(x=depth, y=temperature_C)) +
  ylim(0, 35) +
  geom_point() +
  geom_smooth(method = 'lm', color = "black") +
  labs(
    title = "NTL Temperature by Depth")
print(NTL.tempbydepth)

```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: The plot suggests that the higher the depth (deeper), the lower the temperature of the lake is in celcius. The distribution of points suggests that depth vs temperature has a strong negative correlation. The distribution of points suggests that the relationship between depth vs temperature is relatively linear but the data also suggests that the best goodness of fit line may not be linear. When looking at the depth of 10 or above, the line does not accurately reflect the relationship between the two variables based on the line of best fit. This may suggest that a better goodness of fit between the two variables may be a curved line. There is also a high varianace outside of the line of best fit, which suggests that the correlation between the two variables may not be very high.


7. Perform a linear regression to test the relationship and display the results

```{r linear.regression}
#7

temperature.regression <- lm(data = NTL.LTER.subset, temperature_C ~ depth)
summary(temperature.regression) 

par(mfrow = c(2,2), mar=c(4,4,4,4))
plot(temperature.regression)
par(mfrow = c(1,1))
```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: 73.87% of the variability in temperature is explained by changes in depth. The degrees of freedom is 9,726. The result is statistically significant because the pvalue is 2.2e-16, which is less than 0.05. For a 1m change in depth, the temperature will decrease by 1.94621 degrees celcius. 


---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9
NTL.LTER.AIC <- NTL.LTER.July %>% 
  select(year4, daynum, depth, temperature_C) %>% 
  na.omit(year4, daynum, depth, temperature_C)
  
tempAIC <- lm(data = NTL.LTER.AIC, temperature_C ~ depth + year4 + daynum)
step(tempAIC)
#the set of explanatory variables that is best suited 
#to predict temperature is one that includes all three of those variables. 
#The AIC is 26066.

#10
temp.multipleregression <- lm(data = NTL.LTER.July, temperature_C ~ depth + year4 + daynum)
summary(temp.multipleregression)

```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: The final set of explanatory variables that the AIC method suggests to predict temperature in our multiple regression are year4, daynum, and depth. 74.12% of the variability can be explained by these variables. It is an improvement over the model that only uses depth as an explanatory variable; in that case 73.87% of the variability was explained by just depth. 


---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12
NTL.LTER.July.naomit <- NTL.LTER.July[!(is.na(NTL.LTER.July$temperature_C)), ]
summary(NTL.LTER.July.naomit$temperature_C)

NTL.subset.lakename <- NTL.LTER.July.naomit %>% 
  group_by(lakename, sampledate) %>% 
  summarise(temperature_C = mean(temperature_C))
summary(NTL.subset.lakename)

NTL.subset.lakename.anova <- aov(data = NTL.subset.lakename, temperature_C ~ lakename)
summary(NTL.subset.lakename.anova)

NTL.subset.lakename.anova2 <- lm(data = NTL.subset.lakename, temperature_C ~ lakename)
summary(NTL.subset.lakename.anova2)
```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: The p-value for both the ANOVA and the linear regression are both below 0.05, therefore, it is statistically significant. Therefore, we can reject the null hypothesis and conclude that there is a significant difference in the mean temperatures between the lakes in the month of July. 


14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.

NTL.tempbydepth2 <-
  ggplot(NTL.Lake, aes(x=depth, y=temperature_C, color=lakename)) +
  ylim(0, 35) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = 'lm', se = FALSE, color = "black") +
  labs(
    title = "NTL Temperature by Depth",
    x = "Depth in Meters",
    y = "Temperature in Celcius",
    color = "Lake Name")
print(NTL.tempbydepth2)

#used original dataset because does not specify only July months.
```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15

TukeyHSD(NTL.subset.lakename.anova)
subset.lakename.groups <- HSD.test(NTL.subset.lakename.anova, "lakename", group = TRUE)
subset.lakename.groups

```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: Both Paul Lake and Ward Lake have the same mean temperature, statistically speaking, as Peter Lake. For both Paul Lake and Ward lake have the same mean temperature, statistically speaking, because their p-value is greater than 0.05, so we fail to reject the null hypothesis that the have the same mean temperature. Central Long Lake has a mean temperature that is statistically distinct from all the other lakes because the difference in means of Central Long Lake and all the other lakes has a p-value of less than 0.05. This means that we reject the null hypothesis that the difference in means is 0.


17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: We could also use the two-sample T test because a two-sample T test tests the hypothesis that the two lakes would have the same mean temperature.


18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}
NTL.LTER.specificlakes <- NTL.LTER.July %>% 
  filter(lakename %in% c("Crampton Lake", "Ward Lake"))

Julytemp.twosample <- t.test(NTL.LTER.specificlakes$temperature_C ~ NTL.LTER.specificlakes$lakename)
Julytemp.twosample
```

>Answer: We do not reject the null hypothesis. We cannot conclude that the difference of the mean temperatures for Crampton Lake and Ward Lake is not equal to 0 because the p-value is greater than 0.05 at 0.2649. This supports the results in part 16 because the interaction between the two lakes in the Tukey's HSD test showed the p-value of the difference in means to be greater than 0.05. Therefore, in both tests, we fail to reject the null hypothesis that the difference in means is 0.
